### 2023-03-02 reports/paper/04-discussion.rmd
manuscript prepared

We estimated counterfactual 20-year risks of NHL from 1985 to 2005 in the UAW-GM  Cohort Study under different hypothetical interventions on exposure to soluble MWFs using the hazard-extended ICE parametric g-formula. We found a monotonic exposure-dependent relationship with lower risk estimates arising from stronger limits. Contrasts in estimated NHL risk were statistically significant under the strongest static intervention, which limited soluble MWF exposure to 0.05 mg/m^3^ (a tenth of the current NIOSH REL), and the two dynamic interventions, which sought to limit total MWF exposure to 0.5 and 0.25 mg/m^3^ (the REL and half the REL, respectively) by intervening only on exposure to soluble MWF.

Under the assumptions of conditional exchangeability, positivity, consistency, and correct model specification, our estimates are unbiased for the true counterfactual risk under the hypothetical interventions. Conditional exchangeability means that for all time points, there is no confounding of the relationship between exposure/censoring and NHL status at all future time points given the observed past, including past exposure and covariates. A major threat to conditional exchangeability in longitudinal occupational studies is the HWSE. We limit potential bias due to the HWSE by conditioning on cumulative exposure, employment status, and cumulative time off history at each time point. Cumulative time off and employment status are reasonable mediators of the causal paths linking past health to future exposure and health, but adjustment for these variables may not be sufficient for eliminating bias due to the HWSE. Declines in a worker's health may lead to reductions in work-related exposure without affecting employment status or time off work [@Garcia_2017]. We expect the absence of valid time-varying measures of worker health over the life course to result in bias toward the null.

Positivity refers to the need for adequate variation in future exposure among strata formed by observed covariate and intervention-compliant exposure histories. Even under conditional exchangeability, where exposures within these strata may be considered the result of experimental assignment, expected counterfactual outcomes under different exposures may not be estimable if there is sparsity in the observed distribution of exposures [@Maldonado_2002]. We investigated static and dynamic stochastic interventions on soluble MWF exposure that intervened only when average annual exposure exceeded the hypothetical limit under consideration. Hence, our parameters of interest achieve positivity more easily than those for deterministic interventions e.g. setting all to a single level of exposure. Nonetheless, violations in positivity were still of concern due to the high dimensionality of covariates, as is common in longitudinal settings. We addressed sparsity by summarizing the 20 years of follow-up over a coarser timescale with only 8 follow-up periods and by representing covariates using fewer categories. Coarsening limits the comprehensiveness of confounding control, but improves positivity. In practice, causal inference using observational data must always balance positivity, covariate adjustment, and model specification [@Petersen_2012]. When estimating causal effects using estimators that do not require exposure modeling, as is the case with ICE g-formula methods, concerns over potential bias due to practical violations in positivity may be relaxed if correct model specification, discussed below, is attained.

@@

Correct model specification is a standard assumption in all parametric analyses. The estimator we applied offers greater statistical efficiency than the classical ICE parametric g-formula estimator because it leverages greater parametric smoothing. In causal analyses of longitudinal cohort studies, both the hazard-extended and classical ICE parametric g-formula estimators are less common than the NICE parametric g-formula [@Keil_2014]. However, a major limitation of the NICE g-formula is the g-null paradox: the guaranteed misspecification of parametric models resulting in the false rejection of the null hypothesis when the null is true and when there is time-varying confounding affected by past exposure [@Naimi_2015;@McGrath_2022]. As with all ICE g-formula estimators, the estimator we applied is not subject to the g-null paradox. Furthermore, simulation studies show that the variance of the hazard-extended ICE parametric g-formula is similar to that of the NICE parametric g-formula, so we expect the former to be no less conservative than the latter [@Wen_2020].

@@

Much of the existing epidemiologic literature linking occupational and environmental exposures to NHL report findings from case-control studies where exposures are measured crudely as binary indicators of exposure or membership in a particular occupational group [@Cano_2001;@Costantini_2001;@Karunanayake_2008;@t-Mannetje_2016]. Associations between occupations and NHL risk vary considerably, but one study of working men in Kansas and Nebraska found strong associations between NHL risk and occupations involving metalworking and motor vehicles [@Zheng_2002]. Both of these occupations may entail exposure to soluble MWFs, which contain a number of additives of concern for human health and for NHL risk in particular. Organic compounds containing phosphorous, chlorine, sulfur, nitrogen, and boron are commonly added to soluble MWF to control microbial growth, improve performance under high heat/pressure, and inhibit corrosion [@Evans_2020]. Organophosphorus compounds include organophosphate pesticides, which have been linked to cancer risk in epidemiologic and animal studies. Some were classified as possibly carcinogenic by the IARC [@IARC_2017]. Studies of occupational exposure to chlorinated solvents and pesticides have also been linked to NHL risk [@Cocco_2008;@Purdue_2011;@Cocco_2013;@Vlaanderen_2013;@Callahan_2018]. In 2014, the IARC classified trichloroethylene, tetrachloroethylene, and other chlorinated agents as Group 1 carcinogens [@IARC_2014]. Chlorinated solvents are commonly used as degreasers in industrial settings, but their use in the plants under study here was uncommon and limited to particular operations [@Shrestha_2016]. The structural characteristics shared by MWF additives and known/suspected carcinogens suggest potential similarities in their behavior in biological systems.

This study investigated the effect of hypothetical limits on MWF exposure by comparing the standardized distributions of NHL under various distributions of cumulative exposure induced by applying upper bounds to average annual exposure to soluble MWF. When average annual exposure did not exceed the hypothetical limit, it was left at the observed level of exposure. We selected these hypothetical limits based on the NIOSH REL of 0.5 mg/m^3^ [@niosh_1998]. In the real world, there is no enforcement of the REL, but we nonetheless observed average annual exposures below the REL in these GM plants for many years. If the REL were enforced in the real world, we would not expect reductions in exposure for these low-exposure person-years. Hence, contrasting the counterfactual scenario where all workers experienced average annual exposure at the REL to one where all workers experienced average annual exposure at some higher level would result in an overestimate of the expected real-world benefit of REL enforcement. Furthermore, since the REL is for MWF exposure of all types, we evaluated dynamic stochastic interventions that better reflect this target policy because they reduce exposure to soluble MWF with the goal of reducing MWF exposure of all types to a certain limit. The interventions we considered result in exposure distributions more similar to what we would expect in the real world; our analysis yielded results that more realistically quantify the effects of policy than those investigating deterministic interventions [@Bembom_2007;@Munoz_2012].

>>>>

We estimated counterfactual 20-year risks of NHL from 1985 to 2005 in the UAW-GM  Cohort Study under different hypothetical interventions on exposure to soluble MWFs using the hazard-extended ICE parametric g-formula. We found a monotonic exposure-dependent relationship with lower risk estimates arising from stronger limits.

Under the assumptions of conditional exchangeability, positivity, consistency, and correct model specification, our estimates are unbiased for the true counterfactual risk under the hypothetical interventions. Conditional exchangeability means that for all time points, there is no confounding of the relationship between exposure/censoring and both future exposure/censoring and NHL status given the observed past, including past exposure and covariates [@Richardson_2013;@Diaz_2021]. A major threat to conditional exchangeability in longitudinal occupational studies is the HWSE. We limit potential bias due to the HWSE by conditioning on cumulative exposure, employment status, and cumulative time off history at each time point. Cumulative time off and employment status are reasonable mediators of the causal paths linking past health to future exposure and health, but adjustment for these variables may not be sufficient for eliminating bias due to the HWSE. Declines in a worker's health may lead to reductions in work-related exposure without affecting employment status or time off work [@Garcia_2017]. We expect the absence of valid time-varying measures of worker health over the life course to result in bias toward the null.

Positivity refers to the need for adequate variation in future exposure among strata formed by observed covariate and intervention-compliant exposure histories. Even under conditional exchangeability, where exposures within these strata may be considered the result of experimental assignment, expected counterfactual outcomes under different exposures may not be estimable if there is excessive sparsity in the observed distribution of exposures [@Maldonado_2002]. We investigated static and dynamic stochastic interventions on soluble MWF exposure that intervened only when average annual exposure exceeded the hypothetical limit under consideration. Hence, our parameters of interest achieve positivity more easily than those for static deterministic interventions e.g. setting all to a single level of exposure. Nonetheless, violations in positivity were still of concern due to the high dimensionality of covariates, as is common in longitudinal settings. We addressed sparsity by summarizing the 20 years of follow-up over a coarser timescale with only 8 follow-up periods and by representing covariates using fewer categories. Coarsening limits the comprehensiveness of confounding control, but improves positivity. In practice, causal inference using observational data must always balance positivity, covariate adjustment, and model specification [@Petersen_2012].
<!-- When estimating causal effects using estimators that do not require exposure modeling, as is the case with ICE g-formula methods, concerns over potential bias due to practical violations in positivity may be relaxed if correct model specification, discussed below, is attained. -->

@@

Correct model specification is a standard assumption in all parametric and semi-parametric analyses. The estimator we applied offers greater statistical efficiency than the classical ICE parametric g-formula estimator because it leverages greater parametric smoothing. In causal analyses of longitudinal cohort studies, both the hazard-extended and classical ICE parametric g-formula estimators are less common than the NICE parametric g-formula [@Keil_2014]. However, a major limitation of the NICE g-formula is the g-null paradox: the guaranteed misspecification of parametric models resulting in the false rejection of the null hypothesis when the null is true and when there is time-varying confounding affected by past exposure [@Naimi_2015;@McGrath_2022]. As with all ICE g-formula estimators, the estimator we applied is not subject to the g-null paradox. Furthermore, simulation studies show that the variance of the hazard-extended ICE parametric g-formula is similar to that of the NICE parametric g-formula, so we expect the former to be no less conservative than the latter [@Wen_2020].

@@

Much of the existing epidemiologic literature linking occupational and environmental exposures to NHL report findings from case-control studies where exposures are measured crudely as binary indicators of exposure or membership in a particular occupational group [@Cano_2001;@Costantini_2001;@Karunanayake_2008;@t-Mannetje_2016]. Associations between occupations and NHL risk vary considerably, but one study of working men in Kansas and Nebraska found strong associations between NHL risk and occupations involving metalworking and motor vehicles [@Zheng_2002]. Both of these occupations may entail exposure to soluble MWFs, which contain a number of additives of concern for human health and for NHL risk in particular. Organic compounds containing phosphorous, chlorine, sulfur, nitrogen, and boron are commonly added to soluble MWF to control microbial growth, improve performance under high heat/pressure, and inhibit corrosion [@Evans_2020]. Organophosphorus compounds include organophosphate pesticides, which have been linked to cancer risk in epidemiologic and animal studies. Some were classified as possibly carcinogenic by the IARC [@IARC_2017]. Studies of occupational exposure to chlorinated solvents and pesticides have also been linked to NHL risk [@Cocco_2008;@Purdue_2011;@Cocco_2013;@Vlaanderen_2013;@Callahan_2018]. In 2014, the IARC classified trichloroethylene, tetrachloroethylene, and other chlorinated agents as Group 1 carcinogens [@IARC_2014]. Chlorinated solvents are commonly used as degreasers in industrial settings, but their use in the plants under study here was rare [@Shrestha_2016]. The structural characteristics shared by MWF additives and known/suspected carcinogens suggest potential similarities in their behavior in biological systems.

This study investigated the effect of hypothetical limits on MWF exposure by comparing the standardized distributions of NHL under various distributions of cumulative exposure induced by applying upper bounds to average annual exposure to soluble MWF. We selected these hypothetical limits based on the NIOSH REL of 0.5 mg/m^3^ [@niosh_1998]. In the real world, there is no enforcement of the REL, but we nonetheless observed average annual exposures below the REL in these GM plants for many years. If the REL were enforced in the real world, we would not expect reductions in exposure for these low-exposure person-years. Hence, contrasting the counterfactual scenario where all workers experienced average annual exposure at the REL to one where all workers experienced average annual exposure at some higher level would result in an overestimate of the expected real-world benefit of REL enforcement.

### 2023-03-02 reports/paper/paper.Rmd
manuscript prepared

**Manuscript title:** Evaluating hypothetical limits on metalworking fluid exposure for reducing non-Hodgkin lymphoma incidence: An application of the hazard-extended parametric g-formula

**Authors:** Kevin T. Chen^a,b^, Ellen A. Eisen^b^, Sally Picciotto^b^

@@

@@

**Acknowledgements:**

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased significantly since 1960 and was recently linked with workplace exposure in the United Auto Workers-General Motors (UAW-GM) Cohort Study. Here, we investigate hypothetical interventions on workplace exposure to oil- and water-based soluble metalworking fluids (MWF) in relation to NHL risk. Interventions on entire classes of occupational exposures rather than particular agents may result in stronger lasting protections for worker health.

**Methods:** We estimated counterfactual risk of NHL between 1985 and 2004 in the UAW-GM cohort of autoworkers exposed to straight, soluble, and synthetic MWF at three Michigan plants under hypothetical interventions. We estimated the effect of static and dynamic stochastic interventions on average annual exposure to soluble MWF. Unlike the static interventions, dynamic interventions take into account exposures to the other two classes of metalworking fluids, reflecting the fact that currently recommended limits apply to the total of all three exposures.

**Results:** During follow-up, 231 NHL cases occurred. Stronger hypothetical limits on average annual exposure to soluble MWF resulted in monotonically stronger reductions in NHL risk. Capping exposure to soluble MWF at 0.05 mg/m^3^ resulted in a risk ratio of 0.79 (95% CI: 0.61, 0.96). Capping total MWF exposure at 0.05 mg/m^3^ by intervening only on exposure to soluble MWF resulted in a risk ratio of 0.72 (95% CI: 0.44, 0.97).

**Conclusion:** We considered interventions on soluble MWF that address total MWF exposure levels and showed that stronger limits on average annual exposure to soluble MWF may confer stronger protection against NHL risk.

# What this study adds

(Limit: 100 words)

Little is known about possible links between occupational exposures and non-Hodgkin lymphoma. Existing studies are generally single time-point case-control studies with limited exposure assessment. We leveraged the large study population size and quantitative exposure assessment of the United Auto Workers-General Motors Cohort Study to estimate counterfactual risk of non-Hodgkin lymphoma under realistic interventions using a novel causal estimator capable of adjusting for time-varying confounding affected by past exposure.

>>>>

**Manuscript title:** Hypothetical limits on workplace exposure for reducing non-Hodgkin lymphoma: An application of the hazard-extended iterative conditional expectation parametric g-formula

**Authors:** Kevin T. Chen^a,b,c^, Ellen A. Eisen^b^, Sally Picciotto^b^

@@

^c^ University of California, Berkeley Department of Statistics

@@

**Keywords:** Metalworking fluid, occupational exposure, non-Hodgkin lymphoma, cancer, causal inference, g-formula, healthy-worker effect, United States

@@

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased significantly since 1960 and was recently linked to workplace exposure to soluble metalworking fluids (MWF) in a standard survival analysis of the United Auto Workers-General Motors (UAW-GM) cohort. Here, we investigate the causal effects of hypothetical limits on soluble MWF exposure in relation to NHL risk using an iterative conditional expectation (ICE) parametric g-formula estimator.

**Methods:** We estimated counterfactual risk of NHL between 1985 and 2005 in the UAW-GM cohort of autoworkers at three Michigan plants (n = 34,734) under hypothetical limits on average annual exposure to soluble MWF. We applied the hazard-extended ICE parametric g-formula, an estimator capable of adjusting for time-varying confounding affected by past exposure.

**Results:** During follow-up, 231 NHL cases occurred. Stronger hypothetical limits on soluble MWF exposure resulted in monotonic reductions in NHL risk estimates. Capping average annual exposure at 0.5 mg/m^3^, the NIOSH recommended exposure limit, would have prevented 44 (95% CI: -6, 91) cases. Capping at 0.25 mg/m^3^ would have prevented 52 (95% CI: -5, 106) cases. Finally, capping at 0.05 mg/m^3^ would have prevented 71 (95% CI: 12, 129) cases.

**Conclusion:** Stronger limits on soluble MWF exposure would have prevented more cases of NHL in the UAW-GM cohort. Our application demonstrates the utility of an ICE g-formula estimator for estimating the effect of realistic exposure interventions on a survival outcome. Greater availability of worked examples and software would facilitate wider application of ICE g-formula estimators.
